\documentclass[11pt]{report}

\usepackage{fullpage}

\usepackage{booktabs}

\usepackage{amsmath,amssymb,amstext} % Lots of math symbols and environments
\usepackage[pdftex]{graphicx} % For including graphics N.B. pdftex graphics driver 

%custom packages
\usepackage{cancel}
\usepackage{comment}
\usepackage{courier}
\usepackage{fixltx2e}  % for subscript text in sections/bookmarks
\usepackage{listings}
%\usepackage[12pt]{moresize}
\usepackage{multirow}
%\usepackage{ulem}
\usepackage{url}
\usepackage[utf8]{inputenc}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

\newcommand{\cd}{\ttfamily\footnotesize}

\lstdefinelanguage{ScalaExamples}{
	morekeywords={
		def,var,val,object,package,class,abstract,trait,extends,with,
		match,case,
		import,
		this,super,override,
		private,protected,public,private[this],protected[this],
		final,sealed,null,new,
		for,do,while,until,yield,return},
	sensitive=true,
	morecomment=[l]{//},
	morecomment=[s]{/*}{*/},
	morestring=[b]"
}
\lstset{
	language={ScalaExamples},
	basicstyle=\cd,
	tabsize=2,
	keywordstyle=\bf\cd\color{Blue},
	identifierstyle=\bf\cd,
	commentstyle=\bf\scriptsize\color{gray},
	xleftmargin=\parindent,
	extendedchars=true,
	breaklines=true,
	captionpos=b,
	frame=b,
	float=htp,
	emph={@pure,@readonly,@bounded,@rothis,
		@mutEnv[C],mutEnv[C],@mutEnv[D],mutEnv[D],@mutable,@fresh},
	emphstyle={\bf\cd\color{OliveGreen}},
	moredelim=**[is][\bf\cd\color{OliveGreen}]{@*}{*@},
	escapeinside={(*}{*)}
}

\begin{document}


% all language structures with mutable/readonly
%  0. argument for conceptual simplicity over optimization
%  1. field selection
%  2. method selection
%  3. object construction and type selection
%  4. environment references and closures and package objects
%  5. accessor methods

% extension1: rothis, roparam, fresh

% defaults1: accessors, function types, bounded types, new objects, ...
%   function types and syntactic sugar

% holes1: bounded, asInstanceOf


% basic qualifiers
% qualifiers allowed on fields (vs local variables)

% List of bits to include:
%   rothis - polymorphism
%   viewpoint adaptation
%   closures
%   polymorphic purity
%   fresh results

\title{RIP-Scala:\\Compiler-checkable Reference Immutability and Purity\\for the Scala Language}
\author{Jonathan Rodriguez\\University of Waterloo}
\maketitle

% RESEARCH AGENDA

% does type substitution/refinement create holes in the system?
% under what conditions are methods really pure (not creating externally observable effects)?
% viewpoint adpation and lower bounds of type members
% acceptable qualifier defaults


\chapter{Introduction}

% the motivation behind this work

% background: reference immutability

% RI as a system of type qualifiers

%  lookahead: arbitrary nesting of classes and methods, environment references, closures,
%  env-reference polymorphism, freshness

The motivation behind this work is to help programmers reduce unexpected
mutations of shared state by means of a compile-time annotation-checking mechanism.
Unexpected mutation of shared state is a pervasive problem in object-oriented systems,
both single-threaded and concurrent.
{\em The Geneva Convention on the Treatment of Object Aliasing}~\cite{geneva}
pinned the blame for this problem on the unrestricted copying of object references,
and suggested that such aliasing may be managed through some combination
of aliasing detection, advertisement, prevention, and control.
Of particular interest in the current work are aliasing advertising
and prevention; alias advertising involves statically declaring (for example)
the aliasing operations that a method may perform, and alias prevention involves
the creation of mechanisms for statically checking alias advertisements.

One prominent attack on the problem of object aliasing was Ownership Types~\cite{ownership-types},
which introduced a statically-checkable notion of {\em object ownership},
where references to any given object cannot be created or passed outside of its owning object.
Although the original formulation of ownership types was inflexible
and annotation-heavy, it spawned the research into a large number of
lighter-weight and more-flexible systems of alias restriction.
%increasingly-flexible
%and lighter-weight ownership systems.  % such as universes, lightweight alias protection, etc.
One of the concepts that emerged from the exploration of more flexible systems
is the {\em owners-as-modifiers} discipline, which (unlike the
{\em owners-as-dominators} discipline of the original ownership types system)
allows arbitrary references into an ownership domain provided that
no mutations are performed on objects reached through those references.

{\em Readonly} references turn out to be useful for more than simple support of
ownership type systems.
Even in the absence of explicit aliasing or ownership information,
knowledge of which references are {\em readonly} is frequently sufficient to
prove that certain program methods are pure (or {\em side effect free}).
Where method purity is not guaranteed,
annotations of {\em readonly} references can still serve as compiler-checkable documentation
that certain objects (namely, those reachable through {\em readonly} references) are
not expected to be mutated.
The next chapter (chapter~\ref{chap-background}) discusses the strengths and limitations
of {\em readonly} as a reference annotation mechanism, and how {\em readonly} may be
combined with other types of knowledge to produce stronger guarantees.
Limitations notwithstanding, the addition of {\em readonly} remains a
worthwhile first step in the limitation of shared-memory side effects.

The current work explores a particular approach to introducing {\em readonly}
and related annotations into the Scala language.
Scala includes a number of advanced language-level features that complicate
the practical introduction of reference immutability.
These features include path-dependent type selection, synthetic accessor methods,
and arbitrary nesting of classes and methods.
Scala is not the only programming language with these features;
the ideas in the current work may be transferrable to languages with
similar features, although the investigation of such transferrence is
outside of the scope of the current work.
Correct handling of Scala's features necessitates a more complex approach to
reference immutability than previous approaches for Java,
particularly if the result is to be practical for use in production code.

The focus in the current work is on a compiler-checkable annotation system.
These annotations are compiler-checkable documentation of the ways in which programmers
expect annotated references to be copied and used.
The focus is furthermore on the limitation of shared-memory side effects
rather than the description of general side effects, although further work
may show that reference immutability is sufficient to describe stateful changes
in the external world as well as in memory.
Related dimensions of investigation not addressed by the current work
include run-time considerations, gradual typing considerations,
or whole-program inference.
Although investigation of these dimensions would be useful,
they are outside of the scope of the current work.

The current work defines reference immutability as a system of {\em type qualifiers}.
The type system restricts the set of allowed programs to those programs that
the compiler can prove type-correct; addition of reference-immutability {\em qualifiers}
further restricts allowed programs to those that are consistent with respect to
reference immutability constraints.
A multi-phase compiler plugin implements the necessary reference-immutability-related logic.

Chapter~\ref{chap-background} contains background on reference immutability and
on the Scala language.
Chapter~\ref{chap-mutable-readonly} presents a basic reference immutability system
for the Scala language, and chapter~\ref{chap-flexibility-scala} presents a set of
extensions to the basic system.
Chapter~\ref{chap-dot-calculus-emend} formalizes key parts of the system
as an extension to the DOT calculus.
The type qualifier system itself does not address all practical concerns;
the question of appropriate default qualifiers is discussed in chapter~\ref{chap-defaults},
and mapping of programmer-facing annotations to appropriate qualifiers is
addressed in chaper~\ref{chap-annotations}.
Chapter~\ref{chap-related-work} compares the current work with related work,
and chapter~\ref{chap-evaluation} evalutes the work in terms of support for
code idioms, compiler performance impact, and experience annotating
the standard library.
Future work is briefly described in chapter~\ref{chap-future-work},
and chapter~\ref{chap-conclusion} contains a conclusion.

\chapter{Background} \label{chap-background}
	\input{chap-background}



\chapter{Reference Immutability for Scala}
	\label{chap-mutable-readonly}
	\input{chap-mutable-readonly}


\chapter{Extensions of Reference Immutability for Scala} \label{chap-flexibility-scala}
	\input{chap-flexibility-scala}


\chapter{A Modification of the DOT Calculus} \label{chap-dot-calculus-emend}
	\input{chap-dot-emend}


\chapter{Default Qualifiers} \label{chap-defaults}
	\input{chap-defaults}


\chapter{Programmer-facing Annotations} \label{chap-annotations}
	\input{chap-annotations}

\chapter{Related Work} \label{chap-related-work}
TBD.

\chapter{Evaluation} \label{chap-evaluation}
\section{Code Idioms}
TBD. {\em Iterator pattern, construct immutable circular list.}
\section{Performance \& Memory Usage}
TBD.
\section{Annotating the Standard Library}
TBD.

\chapter{Future Work} \label{chap-future-work}
\section{Observational Exposure}
TBD.

\chapter{Conclusion} \label{chap-conclusion}
TBD.



\bibliographystyle{plain}
\renewcommand*{\bibname}{References}
\bibliography{DotMod2015}
\end{document}